---
title: "Lung: QC, integration, annotation"
author: "GinoBonazza (ginoandrea.bonazza@usz.ch)"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: workflowr::wflow_html
editor_options:
  chunk_output_type: inline
---

```{r knitr config, include=FALSE, cache=FALSE}
knitr::opts_chunk$set(warning = FALSE)

knitr::opts_chunk$set(message = FALSE)

knitr::opts_chunk$set(cache = FALSE)

knitr::opts_chunk$set(dpi = 300, fig.align = "center")
```

## Setup

```{r setup, class.source = "fold-hide"}
# Get current file name to make folder
current_file <- "Lung_QC_integration_annotation"

# Load libraries
library(here)
library(readr)
library(readxl)
library(xlsx)
library(Seurat)
library(DropletUtils)
library(Matrix)
library(scDblFinder)
library(scCustomize)
library(dplyr)
library(ggplot2)
library(magrittr)
library(harmony)
library(tidyverse)
library(reshape2)
library(S4Vectors)
library(SingleCellExperiment)
library(pheatmap)
library(png)
library(gridExtra)
library(knitr)
library(scales)
library(RColorBrewer)
library(Matrix.utils)
library(tibble)
library(ggplot2)
library(scater)
library(patchwork)
library(statmod)
library(clustree)
library(gprofiler2)

#Output paths
output_dir_data <- here::here("output", current_file)
if (!dir.exists(output_dir_data)) dir.create(output_dir_data)

if (!dir.exists(here::here("docs", "figure"))) dir.create(here::here("docs", "figure"))

output_dir_figs <- here::here("docs", "figure", paste0(current_file, ".Rmd"))
if (!dir.exists(output_dir_figs)) dir.create(output_dir_figs)
```

## Quality Control

```{r sample_names}
sample_names <- c("Con1-SN1738",
                  "Con2-SN1846",
                  "Con3-SN1848",
                  "Fosl2tg1-SN1728",
                  "Fosl2tg2-SN1850",
                  "Fosl2tg3-SN1853")

cell_ids <- c("Con1_SN1738",
              "Con2_SN1846",
              "Con3_SN1848",
              "Fosl2tg1_SN1728",
              "Fosl2tg2_SN1850",
              "Fosl2tg3_SN1853")
```

```{r Lung_preQC}
Lung_preQC <- list()

for (sample in sample_names) {
  Lung_preQC[sample] <- Read10X_h5(filename = here::here("data", "Lung", sample, "sample_feature_bc_matrix.h5"))
  Lung_preQC[[sample]] <- CreateSeuratObject(Lung_preQC[[sample]], project = sample)
}

Lung_preQC <- merge(Lung_preQC[[sample_names[1]]], y = c(Lung_preQC[[sample_names[2]]], 
                                                      Lung_preQC[[sample_names[3]]], 
                                                      Lung_preQC[[sample_names[4]]], 
                                                      Lung_preQC[[sample_names[5]]], 
                                                      Lung_preQC[[sample_names[6]]]), 
                    add.cell.ids = c(cell_ids[1], 
                                    cell_ids[2], 
                                    cell_ids[3], 
                                    cell_ids[4], 
                                    cell_ids[5], 
                                    cell_ids[6])
                   )
```

```{r percent.mt}
Lung_preQC$Sample <- Lung_preQC$orig.ident
Lung_preQC[["percent.mt"]] <- PercentageFeatureSet(Lung_preQC, pattern = "^mt-")
Lung_preQC[["percent.rp"]] <- PercentageFeatureSet(Lung_preQC, pattern = "^Rp[sl]")
table(Lung_preQC$Sample)
```

Check quality control parameters

```{r QC_pre_VlnPlots, fig.height=6, fig.width=8}
p1 <- VlnPlot(Lung_preQC, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(Lung_preQC, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(Lung_preQC, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(Lung_preQC, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_VlnPlots
```

```{r QC_pre_VlnPlots_zoom, fig.height=10, fig.width=8}
p1 <- VlnPlot(Lung_preQC, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(Lung_preQC, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 5000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(Lung_preQC, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 25) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(Lung_preQC, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 50) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_VlnPlots_zoom
```

Doublets detection

```{r scDblFinder}
Lung_preQC_sce <- as.SingleCellExperiment(Lung_preQC)
Lung_preQC_sce <- scDblFinder(Lung_preQC_sce, samples="Sample", clusters = TRUE)
table(Lung_preQC_sce@colData$scDblFinder.class)
Lung_preQC <- as.Seurat(Lung_preQC_sce, counts = "counts", data = "logcounts")
rm(Lung_preQC_sce)
Lung_preQC@meta.data[Lung_preQC@meta.data$scDblFinder.class %in% "singlet", "scDblFinder.n"] = paste0("Singlets (n=", table(Lung_preQC$scDblFinder.class)[1], ")")
Lung_preQC@meta.data[Lung_preQC@meta.data$scDblFinder.class %in% "doublet", "scDblFinder.n"] = paste0("Doublets (n=", table(Lung_preQC$scDblFinder.class)[2], ")")
Lung_preQC$scDblFinder.n <- factor(x = Lung_preQC$scDblFinder.n, levels = c(rownames(table(Lung_preQC$scDblFinder.n))[2], rownames(table(Lung_preQC$scDblFinder.n))[1])) 
```

```{r QC_pre_Doublets, fig.height=3, fig.width=8}
p1 <- VlnPlot(Lung_preQC, features = c("nCount_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 50000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(Lung_preQC, features = c("nFeature_RNA"), split.by = "scDblFinder.n", group.by = "Sample", pt.size = 0, y.max = 10000) + theme(axis.title.x = element_blank()) 

QC_pre_Doublets <- p1 + p2 + plot_layout(ncol = 2)

QC_pre_Doublets
```

```{r Save Lung_preQC, eval=FALSE}
saveRDS(Lung_preQC, 
        here::here(output_dir_data, "Lung_preQC.rds"))
```

Remove doublets

```{r Lung_preQC_no_doublets}
Lung_preQC_no_doublets <- subset(x = Lung_preQC, subset = scDblFinder.class == "singlet")
table(Lung_preQC_no_doublets@meta.data$scDblFinder.class)
rm(Lung_preQC)
```

Check quality control parameters after removing doublets

```{r QC_pre_no_doublets_VlnPlots, fig.height=6, fig.width=8}
p1 <- VlnPlot(Lung_preQC_no_doublets, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(Lung_preQC_no_doublets, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(Lung_preQC_no_doublets, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(Lung_preQC_no_doublets, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_no_doublets_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_no_doublets_VlnPlots
```

```{r QC_pre_no_doublets_VlnPlots_zoom, fig.height=10, fig.width=8}
p1 <- VlnPlot(Lung_preQC_no_doublets, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(Lung_preQC_no_doublets, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 5000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(Lung_preQC_no_doublets, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 25) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(Lung_preQC_no_doublets, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 50) + theme(axis.title.x = element_blank()) + NoLegend()

QC_pre_no_doublets_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_pre_no_doublets_VlnPlots_zoom
```

Filter based on number of counts, features and percentage of mitochondrial genes

```{r Lung}
Lung <- subset(Lung_preQC_no_doublets, subset = 
                nFeature_RNA > 200 &
                percent.mt < 25 &
                nCount_RNA > 500)

table(Lung$Sample)
#rm(Lung_preQC_no_doublets)
```

Check quality control parameters

```{r QC_post_VlnPlots, fig.height=6, fig.width=8}
p1 <- VlnPlot(Lung, features = "nCount_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(Lung, features = "nFeature_RNA", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(Lung, features = "percent.mt", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(Lung, features = "percent.rp", group.by = "Sample", pt.size = 0) + theme(axis.title.x = element_blank()) + NoLegend()

QC_post_VlnPlots <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_post_VlnPlots
```

```{r QC_post_VlnPlots_zoom, fig.height=10, fig.width=8}
p1 <- VlnPlot(Lung, features = "nCount_RNA", group.by = "Sample", pt.size = 0, y.max = 20000) + theme(axis.title.x = element_blank()) + NoLegend()

p2 <- VlnPlot(Lung, features = "nFeature_RNA", group.by = "Sample", pt.size = 0, y.max = 5000) + theme(axis.title.x = element_blank()) + NoLegend()

p3 <- VlnPlot(Lung, features = "percent.mt", group.by = "Sample", pt.size = 0, y.max = 25) + theme(axis.title.x = element_blank()) + NoLegend()

p4 <- VlnPlot(Lung, features = "percent.rp", group.by = "Sample", pt.size = 0, y.max = 50) + theme(axis.title.x = element_blank()) + NoLegend()

QC_post_VlnPlots_zoom <- p1 + p2 + p3 + p4 + plot_layout(ncol = 2)

QC_post_VlnPlots_zoom
```

## Integration and clustering

Normalization and scaling.

```{r eval=FALSE}
DefaultAssay(Lung) <- "RNA"
Lung_not_integrated <- Lung %>%
    NormalizeData() %>%
    FindVariableFeatures(selection.method = "vst", nfeatures = 2000) %>% 
    ScaleData(features = rownames(Lung), vars.to.regress = "percent.mt")
```

PCA 

```{r, eval=FALSE, include=FALSE}
Lung_not_integrated <- RunPCA(Lung_not_integrated, npcs = 50)
```

```{r eval=FALSE}
saveRDS(Lung_not_integrated, 
        here::here(output_dir_data, "Lung_not_integrated.rds"))
```

```{r}
Lung_not_integrated <- readRDS(here::here(output_dir_data, "Lung_not_integrated.rds"))
```


```{r Elbow_not_integrated, fig.width=4, fig.height=4}
ElbowPlot(Lung_not_integrated, ndims = 50)
```

Clustering without integration

```{r eval=FALSE}
Lung_not_integrated <- RunUMAP(Lung_not_integrated, dims = 1:40)
Lung_not_integrated <- FindNeighbors(Lung_not_integrated, dims = 1:40)
Lung_not_integrated <- FindClusters(Lung_not_integrated, resolution = seq(0.1, 0.8, by=0.1))
```

```{r Clustree_not_integrated, fig.width=6, fig.height=9}
clustree::clustree(Lung_not_integrated@meta.data[,grep("RNA_snn_res", colnames(Lung_not_integrated@meta.data))],
                   prefix = "RNA_snn_res.")
```

```{r Add_group_info}
Lung_not_integrated@meta.data[Lung_not_integrated@meta.data$Sample %in% c("Con1-SN1738", "Con2-SN1846", "Con3-SN1848"), "Group"] = "WT"
Lung_not_integrated@meta.data[Lung_not_integrated@meta.data$Sample %in% c("Fosl2tg1-SN1728", "Fosl2tg2-SN1850", "Fosl2tg3-SN1853"), "Group"] = "Fosl2tg"
```

```{r UMAPs_not_integrated, fig.width=14, fig.height=4}
DimPlot(Lung_not_integrated, reduction = "umap", shuffle = T,
        group.by = c("RNA_snn_res.0.2", "Sample", "Group"), ncol = 3)
```

```{r eval=FALSE}
saveRDS(Lung_not_integrated, 
        here::here(output_dir_data, "Lung_not_integrated.rds"))
```

Integrate the samples using Harmony

```{r eval=FALSE}
Lung_integrated <- RunHarmony(Lung_not_integrated, assay.use="RNA", group.by.vars = "Sample", dims.use = 1:50)
```

```{r eval=FALSE}
Lung_integrated <- RunUMAP(Lung_integrated, dims = 1:40, reduction = "harmony")
Lung_integrated <- FindNeighbors(Lung_integrated, dims = 1:40, reduction = "harmony")
Lung_integrated <- FindClusters(Lung_integrated, resolution = seq(0.1, 0.8, by=0.1))
```

```{r eval=FALSE}
saveRDS(Lung_integrated, 
        here::here(output_dir_data, "Lung_integrated.rds"))
```

```{r}
Lung_integrated <- readRDS(here::here(output_dir_data, "Lung_integrated.rds"))
```

```{r Clustree_integrated, fig.width=6, fig.height=9}
clustree::clustree(Lung_integrated@meta.data[,grep("RNA_snn_res", colnames(Lung_integrated@meta.data))],
                   prefix = "RNA_snn_res.")
```

```{r UMAPs_integrated, fig.width=14, fig.height=4}
DimPlot(Lung_integrated, reduction = "umap", shuffle = T,
        group.by = c("RNA_snn_res.0.2", "Sample", "Group"), ncol = 3, label = T)
```

Find the markers that characterize each cell population

```{r}
DefaultAssay(Lung_integrated) <- "RNA"
Idents(Lung_integrated) <- "RNA_snn_res.0.2"
Markers <- FindAllMarkers(Lung_integrated, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.5)
write.csv(Markers, here::here(output_dir_data, "Lung_Markers_all.csv"))
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "Lung_Markers_top10.csv"))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "Lung_Markers_top3.csv"))
```

```{r Heatmap_low_quality, fig.width=15, fig.height=10}
mapal <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
mapal <- rev(mapal[1:256])
Heatmap <- DoHeatmap(Lung_integrated, draw.line = F, features = Markers_top10$gene) +
  scale_fill_gradientn(colours = mapal) +
  theme(text = element_text(size = 15), axis.text.y = element_text(size = 5)) +
  theme(plot.margin = unit(c(0.1, 0, 0, 0), 
                           "inches"))
Heatmap
```

Check QC parameters in each cluster

```{r QC_VlnPlots_low_quality, fig.width=10, fig.height=6}
VlnPlot(Lung_integrated, features = c("nCount_RNA", "nFeature_RNA", "percent.mt", "percent.rp"), ncol = 2, pt.size = 0)
```

Check cell cycle genes

```{r Cell_cycle_UMAP, fig.width=5, fig.height=4}
DefaultAssay(Lung_integrated) <- "RNA"
mmus_s = gorth(cc.genes.updated.2019$s.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
mmus_g2m = gorth(cc.genes.updated.2019$g2m.genes, source_organism = "hsapiens", target_organism = "mmusculus")$ortholog_name
Lung_integrated <- CellCycleScoring(Lung_integrated, s.features = mmus_s, g2m.features = mmus_g2m)
DimPlot(Lung_integrated, reduction = "umap", shuffle = T,
        group.by = "Phase")
```


## Cell type annotation

Annotate the clusters based on the characteristic markers

```{r}
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("0"), "cell_type"] = "Endothelial cells"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("1"), "cell_type"] = "T cells"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("2"), "cell_type"] = "Alveolar macrophages"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("3"), "cell_type"] = "B cells"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("4"), "cell_type"] = "Monocytes"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("5"), "cell_type"] = "Fibroblasts"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("6"), "cell_type"] = "Neutrophils"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("7"), "cell_type"] = "AT1/AT2 cells"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("8"), "cell_type"] = "Dendritic cells"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("9"), "cell_type"] = "Interstitial macrophages"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("10"), "cell_type"] = "Aerocytes"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("11"), "cell_type"] = "Red blod cells"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("12"), "cell_type"] = "Proliferating cells"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("13"), "cell_type"] = "Club/Ciliated/Basal cells"
Lung_integrated@meta.data[Lung_integrated@meta.data$RNA_snn_res.0.2 %in% c("14"), "cell_type"] = "Lymphatic cells"

Lung_integrated$cell_type <- factor(Lung_integrated$cell_type, levels = c("Endothelial cells", 
                                                                        "T cells",
                                                                        "Alveolar macrophages",
                                                                        "B cells",
                                                                        "Monocytes",
                                                                        "Fibroblasts",
                                                                        "Neutrophils",
                                                                        "AT1/AT2 cells",
                                                                        "Dendritic cells",
                                                                        "Interstitial macrophages",
                                                                        "Aerocytes",
                                                                        "Red blod cells",
                                                                        "Proliferating cells",
                                                                        "Club/Ciliated/Basal cells",
                                                                        "Lymphatic cells"))
Idents(Lung_integrated) <- Lung_integrated$cell_type
```

```{r FindAllMarkers cell_type}
DefaultAssay(Lung_integrated) <- "RNA"
Idents(Lung_integrated) <- "cell_type"
Markers <- FindAllMarkers(Lung_integrated, only.pos = TRUE, min.pct = 0.5, logfc.threshold = 0.5)
write.csv(Markers, here::here(output_dir_data, "Lung_Markers_all_cell_type.csv"))
Markers_top10 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 10, wt = avg_log2FC))
write.csv(Markers_top10, here::here(output_dir_data, "Lung_Markers_top10_cell_type.csv"))
Markers_top3 <- as.data.frame(Markers %>% group_by(cluster) %>% top_n(n = 3, wt = avg_log2FC))
write.csv(Markers_top3, here::here(output_dir_data, "Lung_Markers_top3_cell_type.csv"))
```

```{r Heatmap, fig.width=12, fig.height=8}
mapal <- colorRampPalette(RColorBrewer::brewer.pal(9,"RdBu"))(256)
mapal <- rev(mapal[1:256])
Heatmap <- DoHeatmap(Lung_integrated, label = FALSE, draw.line = F, features = Markers_top10$gene) + 
  scale_fill_gradientn(colours = mapal) +
  theme(text = element_text(size = 12), 
        axis.text.y = element_text(size = 4.5)) +
  theme(plot.margin = unit(c(0, 0, 0, 0), "inches")) +
  guides(fill = guide_colorbar(barwidth = 1, barheight = 4))
Heatmap
```

```{r UMAP_cell_type, fig.width=6.5, fig.height=6}
p <- DimPlot(Lung_integrated, group.by = "cell_type", reduction = "umap", label = F, shuffle = T) + 
  NoLegend() + 
  theme(axis.text=element_text(size=16, face = "bold"), axis.title = element_text(size = 20, face = "bold")) + 
  theme(plot.title = element_blank())
LabelClusters(p, id = "cell_type", fontface = "bold", size = 4, repel = T)
```

```{r DotPlot_markers, fig.width=8, fig.height=12}
DotPlot(Lung_integrated, assay = "RNA", features = rev(Markers_top3$gene), dot.scale = 5, cluster.idents = FALSE) +
  RotatedAxis() +
  coord_flip() +
  theme(axis.title = element_blank(), axis.text.x = element_text(size = 14), axis.text.y = element_text(size = 13), legend.text = element_text(size = 9), legend.title = element_text(size = 11), plot.margin = unit(c(0, 0, 0, 0.1), 
                           "inches"))
```

```{r FeatPlot_markers, fig.width=20, fig.height=10}
FeaturePlot(Lung_integrated, features = c("Ptprb", "Cd28", "Lpl", "Ms4a1", "Ly6c2", "Col1a2", "S100a8", "Sftpc", "Ccl17", "C1qa", "Ednrb","Hba-a1", "Top2a", "Cyp2f2", "Mmrn1", "Lyz2", "Nkg7", "Ager", "Aqp5", "Itgax", "Ccr2", "Plac8", "Cd163", "Mrc1", "Scgb1a1", "Foxj1", "Muc5ac", "Krt5"), ncol = 7)
```

```{r save Lung_integrated final, eval = FALSE}
saveRDS(Lung_integrated, 
        here::here(output_dir_data, "Lung_integrated.rds"))
```
